Flatiron cleaning

### 2021-12-01 Rmd/rdi analysis.Rmd
Reorganize steps of cleaning and filtering for step 1 and 2

***
At this point, after cleaning the clinical data and excluding unfit patients, we have `r nrow(clinical_data)` patients.  
- exclude the patients who had surgery but we don't have a date  
- exclude patients whose histology is unknown  
- exclude patients who didn't survive 1 month

Now drug cleaning!

# I. Drug episode questions

**We start with a number of patient who have drug info of `r drugs %>% distinct(patientid) %>% nrow()`.**

### step 1: Patients with a date of drug or date of surgery BEFORE diagnosis will be excluded
```{r drugs episode cleanup}
drugs1 <- drugs %>% 
  filter(episodedatasource == "Administrations" #& 
           # ismaintenancetherapy == "FALSE"
         ) %>%
  # 1. 60 days rule between diagnosis and treatment # Loss 1
  # bind with clinical
  inner_join(., clinical_data, 
          by = "patientid") %>% 
  arrange(patientid, episodedate) %>% 
  group_by(patientid) %>% 
  mutate(first_drug_date = first(episodedate)) %>% 
  mutate(days_at_first_drug = interval(start = diagnosisdate, end = first_drug_date)/
           duration(n=1, units = "months")) %>% 
  mutate(days_at_surg = interval(start = diagnosisdate, end = surgerydate)/
           duration(n=1, units = "months")) %>% 
  ungroup()
```
We have `r drugs1 %>% filter(days_at_first_drug < 0) %>% distinct(patientid) %>% nrow()` patients with date of drug before diagnosis and `r drugs1 %>% filter(days_at_surg < 0) %>% distinct(patientid) %>% nrow()` patients with a date of surgery before diagnosis. Patients left :

```{r}
drugs1 <- drugs1 %>% 
  filter(days_at_first_drug >= 0 &
           (is.na(surgerydate) | days_at_surg >= 0)
         )
drugs1 %>% distinct(patientid) %>% nrow()
```

### step 2: Flatiron recommendation cleanup
#### 90 days rule between diagnosis and treatment

Flatiron propose a conceptual instruction on the 90-day-gap : 90 days between diagnosis and treatment.  
It is meant to exclude patients who didn't get treatment for 90 days which would most likely mean that some treatment data are missing. Patients left after this step:
```{r}
remove_90daysrule <- drugs1 %>% 
  filter((days_at_first_drug > 90 | days_at_surg > 90)) %>% 
  distinct(patientid)
remove_90daysrule <- paste(remove_90daysrule$patientid, collapse = "|")

drugs1 <- drugs1 %>% 
  # remove these patients who 90 days rule
  filter(!str_detect(patientid, remove_90daysrule))

drugs1 %>% distinct(patientid) %>% nrow()
```

#### 30 days rule excluding patient who didn’t survive for at least 30 days after diagnosis

This is something that Flatiron strongly recommends, but does not require. Flatiron proposal reviewers will likely provide comments on this if not incorporated into a proposal. 
```{r}
# drugs1 %>% 
#   mutate(thirty_days_exclusion = interval(start = diagnosisdate, end = dateofdeath)/
#            duration(n=1, units = "days"), 
#          thirty_days_exclusion = case_when(
#            thirty_days_exclusion <= 30                  ~ "Exclude"
#          )) %>% 
#   filter(thirty_days_exclusion == "Exclude") %>% 
#   distinct(patientid) %>% 
#   nrow

drugs1 <- drugs1 %>% 
  mutate(thirty_days_exclusion = interval(start = diagnosisdate, end = dateofdeath)/
           duration(n=1, units = "days"), 
         thirty_days_exclusion = case_when(
           thirty_days_exclusion <= 30                  ~ "Exclude",
           thirty_days_exclusion > 30                   ~ "Include",
           is.na(dateofdeath)                           ~ "Include"
         )) %>% 
  filter(thirty_days_exclusion != "Exclude")

# drugs1 %>% distinct(patientid) %>% nrow()

```
In the data, `r drugs1 %>% mutate(thirty_days_exclusion = interval(start = diagnosisdate, end = dateofdeath)/duration(n=1, units = "days"), thirty_days_exclusion = case_when(thirty_days_exclusion <= 30~ "Exclude")) %>% filter(thirty_days_exclusion == "Exclude") %>% distinct(patientid) %>% nrow` are excluded so we still have `r drugs1 %>% distinct(patientid) %>% nrow()` patients.

#### Exclude exclude == "1" coded by Lauren

Are excluded :  
-	Borderline cases (not invasive ovarian cancer)  
-	Data from one PracticeID  (Flatiron requested removal of this practice ID from the dataset)  
-	Certain historical structured data were not available for patients seen at academic sites prior to 2015 – Flatiron has asked that if conducting survival analyses, these patients are removed (patients diagnosed < 2015 at academic sites)  
Patients left after this step:

```{r}
drugs1 <- drugs1 %>% 
  filter(exclude != "1")

drugs1 %>% distinct(patientid) %>% nrow()
```

<!-- ### step .: Remove linename that doesn't include platin or taxanes in first line -> later in the process, need first to recode linenumber with all the drugs in. -->
```{r}
# drugs1 <- drugs1 %>% 
#   # 2. Filtering----
#   filter(episodedatasource == "Administrations" & 
#            # ismaintenancetherapy == "FALSE" & 
#            !is.na(amount) &
#            (days_at_first_drug < 60 | days_at_surg < 60) &
#            
#            days_at_first_drug >= 0 &
#            (is.na(surgerydate) | days_at_surg >= 0)
#          )

# drugs1 %>% 
#   filter(linenumber == 1) %>% 
#   distinct(patientid, .keep_all = TRUE) %>% 
#   select(linename) %>% 
#   tbl_summary(sort = list(everything() ~ "frequency"))
```

### step 3: Remove chemotherapy only patients
```{r}
drugs1 <- drugs1 %>%
  filter(!is.na(surgerydate))
drugs1 %>% distinct(patientid) %>% nrow()
```

@@

### step 5: remove patients with maintenance as first line adjuvant?????
`r emo::ji("question_mark")`
<span style="color:red">**Patients have maintenance as first line Adjuvant then recurrent treatment : Should I remove these patients or keep them for the Neoadjuvant RDI??**</span>
```{r}
therapy_line %>% 
  filter(linenumber_adj == 1) %>% 
  distinct(patientid, linenumber_new, .keep_all = TRUE) %>% 
  select(ismaintenancetherapy, linename) %>% 
  tbl_summary(by = linename)
therapy_line %>% 
  filter(linenumber_adj == 1) %>%
  distinct(patientid, linenumber_new, .keep_all = TRUE) %>% 
  select(ismaintenancetherapy, chemotherapy_type) %>% 
  tbl_summary(by = chemotherapy_type)
therapy_line %>% 
  filter(linenumber_adj == 1, chemotherapy_type == "Neoadjuvant") %>%
  arrange(desc(ismaintenancetherapy)) %>% 
  distinct(patientid, linenumber_new, .keep_all = TRUE) %>% 
  select(ismaintenancetherapy, chemotherapy_type) %>% 
  tbl_summary(by = chemotherapy_type)
```

@@

For these patients, I was worried that the part of their first line which was switched to Adjuvant line because of surgery is short... How long is this adjuvant line?
```{r}
therapy_line %>% 
  mutate(line_duration = interval(start = linestartdate_new, end = lineenddate_new)/ 
           duration(n = 1, units = "days")) %>% 
  filter(linenumber == 1 & linenumber_new == 2) %>% 
  # select(patientid, linenumber, linenumber_new, linenumber_adj, 
  #        drugname, chemotherapy_type, therapy, episodedate, line_duration) %>% 
  distinct(patientid, .keep_all = TRUE) %>% 
  ggplot(aes(x= line_duration))+
  geom_histogram(color = "slateblue", aes(fill = route), binwidth = 7) +
  theme_minimal(base_size = 14)+
  labs(x="Line duration", 
       y="Number of Patient",
       caption = "Each bar represents 7 days, bottom plot is zoomed in")+
  ggforce::facet_zoom(xlim = c(0, 100), zoom.size = 1)
```
Several patients have a short line.  
`r emo::ji("question_mark")` **Should me remove the patient with a line during less than a cycle** We said during October meeting to check at IP

@@

### step 9: Remove too much cycle
```{r more than 9 cycle patients}
therapy_cycle %>% 
  filter(therapy == "Upfront Chemo") %>% 
  mutate(cycle_increment = factor(cycle_increment)) %>% 
  arrange(desc(cycle_increment)) %>%
  distinct(patientid, .keep_all = TRUE) %>%
  select(cycle_increment, chemotherapy_type) %>% 
  tbl_summary(by = chemotherapy_type)
therapy_cycle %>% 
  filter(therapy == "Upfront Surgery") %>% 
  mutate(cycle_increment = factor(cycle_increment)) %>% 
  arrange(desc(cycle_increment)) %>%
  distinct(patientid, .keep_all = TRUE) %>%
  select(cycle_increment, chemotherapy_type) %>% 
  tbl_summary(by = chemotherapy_type)
```
We said to remove patients with more than 9 total cycle but maybe we can save some....

@@

```{r ggridges, fig.width=16, fig.height=18}
library(ggridges)
Frontline %>% filter(!is.na(mean_RDI_per_drug)) %>%
  ggplot(aes(y=raceeth, x=mean_RDI_per_drug,  fill=raceeth
             )) +
    geom_density_ridges(alpha=0.5) +
    theme_ridges()+
  theme(text = element_text(size = 14),#axis.text.y = element_blank(),
        panel.spacing = unit(0.1, "lines"),
        strip.text.x = element_text(size = 8)
    ) +
  geom_vline(xintercept = 1, color = "red", linetype = 2)+

@@

ggsurvplot(survfit(Surv(month_at_os_from_treatment, vitalstatus) ~ RDI_grp, 
                   data=Frontline %>% 
                     filter(drugname == "paclitaxel" & 
                              therapy == "Upfront Chemo") %>%
                     distinct(patientid, .keep_all = TRUE)
                   ),
           title = "OS from first treatment pacliatxel \nUpfront Chemo",
           font.main = c(24, "bold", "black"),
           font.x = c(20, "bold", "black"), font.y = c(20, "bold", "black"),
           font.legend = c(14, "bold", "black"), font.tickslab = c(18, "bold", "black"),
           size = 1.5,

@@

```{r echo=TRUE}
# Highly underdosed: < 0.75
# Underdosed: >= 0.75 to <0.85
# Adequate dosing: >= 0.85 to <1.15
# Overdosing: >=1.15
```
Fun to see the dots for underdosed patients being located on the right side (higher BMI) and dots for overdosed located on the left (lower BMI)
```{r, fig.width=12, fig.height=12}
Frontline %>% filter(!is.na(mean_RDI_per_drug)) %>%
  ggplot(aes(x= bmi_at_dx, y = mean_RDI_per_drug, color= RDI_grp1))+
  geom_point(alpha = 0.7, size = 0.75)+
  theme_minimal(base_size = 10)+
  facet_wrap(drugname~ therapy, ncol = 2, scales = "free")
Frontline %>% filter(!is.na(mean_RDI_per_drug)) %>%
  ggplot(aes(x= bmi_cat, y = mean_RDI_per_drug, color= RDI_grp1))+
  geom_jitter()+
  theme_minimal(base_size = 10)+
  facet_wrap(drugname~ therapy, ncol = 2, scales = "free")

>>>>

# I. Drug episode cleaning

```{r drugs episode cleanup}
drugs <- drugs %>% 
  filter(episodedatasource == "Administrations" #& 
           # ismaintenancetherapy == "FALSE" # keep maintenance for a correct recoding of line number
         ) %>%
  # bind with clinical
  inner_join(., clinical_data, 
          by = "patientid") %>% 
  arrange(patientid, episodedate)
```

**We start with a number of `r drugs %>% distinct(patientid) %>% nrow()` patients who received treatment and for who their clinical info are available.**

## 1. step 1: Flatiron recommendation cleanup

This step is to remove patients from the population that don't follow flatiron recommendations.

- 90 days rule between diagnosis and treatment  
Flatiron propose a conceptual instruction on the 90-day-gap : 90 days between diagnosis and treatment.  
It is meant to exclude patients who didn't get treatment for 90 days which would most likely mean that some treatment data are missing. Patients left after this step:

- 30 days rule excluding patient who didn’t survive for at least 30 days after diagnosis  
This is something that Flatiron strongly recommends, but does not require. Flatiron proposal reviewers will likely provide comments on this if not incorporated into a proposal.

- Exclude exclude == "1" coded by Lauren

Are excluded :  
-	Borderline cases (not invasive ovarian cancer)  
-	Data from one PracticeID  (Flatiron requested removal of this practice ID from the dataset)  
-	Certain historical structured data were not available for patients seen at academic sites prior to 2015 – Flatiron has asked that if conducting survival analyses, these patients are removed (patients diagnosed < 2015 at academic sites)

```{r rule between diagnosis and treatment}
drugs1 <- drugs %>% 
  # 1. 60 days rule between diagnosis and treatment # Loss 1
  group_by(patientid) %>% 
  mutate(first_drug_date = first(episodedate)) %>% 
  mutate(days_at_first_drug = interval(start = diagnosisdate, end = first_drug_date)/
           duration(n=1, units = "months")) %>% 
  mutate(days_at_surg = interval(start = diagnosisdate, end = surgerydate)/
           duration(n=1, units = "months")) %>% 
  ungroup() %>% 
  mutate(thirty_days_exclusion = interval(start = diagnosisdate, end = dateofdeath)/
           duration(n=1, units = "days")
  )
```

```{r}
drugs1 <- drugs1 %>% 
  mutate(flatiron_rules = case_when(
    days_at_first_drug > 90 | 
      days_at_surg > 90                          ~ "exclude",
    thirty_days_exclusion <= 30                  ~ "exclude",
    exclude == "1"                               ~ "exclude"
  )) %>%
  group_by(patientid) %>% 
  fill(flatiron_rules, .direction = "downup")

drugs1 <- drugs1 %>% 
  filter(is.na(flatiron_rules))
```

**At the end of the step 1, we have `r drugs1 %>% distinct(patientid) %>% nrow()` patients.**

## 2. step 2: Filter patients with cinical characteristics fitting to the study

- exclude patients whose histology is unknown  
- remove chemotherapy only patients
```{r}
drugs1 <- drugs1 %>% 
  filter(!is.na(histology)) # loss 287 ################## Will do when bind to drugs
```

```{r}
drugs1 <- drugs1 %>%
  filter(!is.na(surgerydate))
```

**At the end of the step 2, we have `r drugs1 %>% distinct(patientid) %>% nrow()` patients.**

## 3. step 3: Filter out patients with missing data

- exclude the patients who had surgery but we don't have a date of surgery

```{r}
drugs1 <- drugs1 %>% 
  filter(!(issurgery == "Yes" & is.na(surgerydate))) 
```

### step 1: Patients with a date of drug or date of surgery BEFORE diagnosis will be excluded (inclined to be wrongly recorded)
We have `r drugs1 %>% filter(days_at_first_drug < 0) %>% distinct(patientid) %>% nrow()` patients with date of drug before diagnosis and `r drugs1 %>% filter(days_at_surg < 0) %>% distinct(patientid) %>% nrow()` patients with a date of surgery before diagnosis. Patients left :
```{r}
drugs1 <- drugs1 %>% 
  filter(days_at_first_drug >= 0 &
           (is.na(surgerydate) | days_at_surg >= 0)
         )
drugs1 %>% distinct(patientid) %>% nrow()
```

@@

### step 5: remove patients with maintenance as first line adjuvant?????
`r emo::ji("question_mark")`
<span style="color:red">**Patients have maintenance as first line Adjuvant then recurrent treatment : Should I remove these patients or keep them for the Neoadjuvant RDI??**</span>
```{r}
therapy_line %>% 
  filter(linenumber_adj == 1) %>% 
  distinct(patientid, linenumber_new, .keep_all = TRUE) %>% 
  select(ismaintenancetherapy, linename) %>% 
  tbl_summary(by = linename)
# a <- therapy_line %>% 
#   filter(linenumber_adj == 1)
therapy_line %>% 
  filter(linenumber_adj == 1) %>%
  distinct(patientid, linenumber_new, .keep_all = TRUE) %>% 
  select(ismaintenancetherapy, chemotherapy_type) %>% 
  tbl_summary(by = chemotherapy_type)
therapy_line %>% 
  filter(linenumber_adj == 1, chemotherapy_type == "Neoadjuvant") %>%
  arrange(desc(ismaintenancetherapy)) %>% 
  distinct(patientid, linenumber_new, .keep_all = TRUE) %>% 
  select(ismaintenancetherapy, chemotherapy_type) %>% 
  tbl_summary(by = chemotherapy_type)
```

@@

For these patients, I was worried that the part of their first line which was switched to Adjuvant line because of surgery is short... How long is this adjuvant line?
```{r}
therapy_line %>% 
  mutate(line_duration = interval(start = linestartdate_new, end = lineenddate_new)/ 
           duration(n = 1, units = "days")) %>% 
  filter(linenumber == 1 & linenumber_new == 2) %>% 
  # select(patientid, linenumber, linenumber_new, linenumber_adj, 
  #        drugname, chemotherapy_type, therapy, episodedate, line_duration) %>% 
  distinct(patientid, .keep_all = TRUE) %>% 
  ggplot(aes(x= line_duration))+
  geom_histogram(aes(fill = route), binwidth = 7) +
  theme_minimal(base_size = 14)+
  labs(x="Line duration", 
       y="Number of Patient",
       caption = "Each bar represents 7 days, bottom plot is zoomed in")+
  ggforce::facet_zoom(xlim = c(0, 100), zoom.size = 1)
```
Several patients have a short line.  
`r emo::ji("question_mark")` **Should we remove the patient with a line lasting less than a cycle** We said during October meeting to check if were IP administration....

@@

### step 9: Remove Patients with too many cycles
```{r more than 9 cycle patients}
therapy_cycle %>% 
  filter(therapy == "Upfront Chemo") %>% 
  mutate(cycle_increment = factor(cycle_increment)) %>% 
  arrange(desc(cycle_increment)) %>%
  distinct(patientid, .keep_all = TRUE) %>%
  select(cycle_increment, chemotherapy_type) %>% 
  tbl_summary(by = chemotherapy_type)
therapy_cycle %>% 
  filter(therapy == "Upfront Surgery") %>% 
  mutate(cycle_increment = factor(cycle_increment)) %>% 
  arrange(desc(cycle_increment)) %>%
  distinct(patientid, .keep_all = TRUE) %>%
  select(cycle_increment, chemotherapy_type) %>% 
  tbl_summary(by = chemotherapy_type)
```
We said to remove patients with more than 9 total cycle but maybe we can save some....

@@

```{r ggridges, fig.width=16, fig.height=18}
library(ggridges)
Frontline %>% filter(!is.na(mean_RDI_per_drug)) %>%
  ggplot(aes(y=raceeth, x=mean_RDI_per_drug,  fill=raceeth
             )) +
    geom_density_ridges(alpha=0.5) +
    theme_ridges()+
  theme(#text = element_text(size = 14),#axis.text.y = element_blank(),
        panel.spacing = unit(0.1, "lines"),
        strip.text.x = element_text(size = 14)
    ) +
  geom_vline(xintercept = 1, color = "red", linetype = 2)+

@@

ggsurvplot(survfit(Surv(month_at_os_from_treatment, vitalstatus) ~ RDI_grp, 
                   data=Frontline %>% 
                     filter(drugname == "paclitaxel" & 
                              therapy == "Upfront Chemo") %>%
                     distinct(patientid, .keep_all = TRUE)
                   ),
           title = "OS from first treatment paclitaxel \nUpfront Chemo",
           font.main = c(24, "bold", "black"),
           font.x = c(20, "bold", "black"), font.y = c(20, "bold", "black"),
           font.legend = c(14, "bold", "black"), font.tickslab = c(18, "bold", "black"),
           size = 1.5,

@@

```{r echo=TRUE}
# Highly underdosed: < 0.75
# Underdosed: >= 0.75 to <0.85
# Adequate dosing: >= 0.85 to <1.10
# Overdosing: >=1.10
```
Fun to see the dots for underdosed patients being located on the right side (higher BMI) and dots for overdosed located on the left (lower BMI)
```{r, fig.width=12, fig.height=12}
Frontline %>% filter(!is.na(mean_RDI_per_drug)) %>%
  ggplot(aes(x= bmi_at_dx, y = mean_RDI_per_drug, color= RDI_grp1))+
  geom_point(alpha = 0.7, size = 0.75)+
  theme_minimal(base_size = 10)+
  facet_wrap(drugname~ therapy, ncol = 2, scales = "free")
Frontline %>% filter(!is.na(mean_RDI_per_drug)) %>%
  ggplot(aes(x= bmi_cat, y = mean_RDI_per_drug, color= RDI_grp1))+
  geom_jitter()+
  theme_minimal(base_size = 10)+
  facet_wrap(drugname~ therapy, ncol = 2, scales = "free")
